---
title: "03_cluster_umap_analysis"
author: "Caroline_Duncombe"
date: "2026-01-07"
output: html_document
---

# CLuster Analyis and UMAP Generation

Script will take the normalized/batch corrected data, integrate, and produce UMAPS for the manuscript. 

==============================================================================
SETUP AND FILE PATHS
==============================================================================

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# Load libraries
library(dplyr)
library(Seurat)
library(cluster)
library(ComplexHeatmap)
library(tibble)
library(readr)
library(mclust)
library(ggplot2)
library(cowplot)
library(reshape2)
library(RColorBrewer)
library(scales)
library(patchwork)

# Clear environment
rm(list = ls())

library(here)

# here() automatically detects your project root, but you can override it:
here::set_here("/Users/carduncombe/Users/carduncombe/Documents/GitHub/Duncombe-et-al-Front-Immunol-2026")

# Define file paths using here()
input_dir <- here("clean_data", "srt")
output_dir_figs <- here("figs", "UMAPS")
output_dir_data <- here("clean_data", "UMAPS")

# Create output directories if they don't exist
dir.create(output_dir_figs, recursive = TRUE, showWarnings = FALSE)
dir.create(output_dir_data, recursive = TRUE, showWarnings = FALSE)

# Define input files using here()
input_seurat <- here("clean_data", "srt", "QC_normalized_data_200um_SVA.rds")
output_seurat_cca <- here("clean_data", "UMAPS", "QC_normalized_data_200um_SVA_CCA.rds")

# Load data
dat <- readRDS(input_seurat)
```

==============================================================================
STEP 1: QUALITY CHECK - SVA BATCH CORRECTION
==============================================================================

```{r step 1}
# Extract SVA batch-corrected counts
svacounts <- GetAssayData(dat, assay = "SVA_batchCorrected", layer = "counts")

# Extract the original GeoMx counts
geomx_counts <- GetAssayData(dat, assay = "GeoMx", layer = "counts")

# Store the summed counts as metadata
dat_test <- AddMetaData(dat, metadata = Matrix::colSums(svacounts), col.name = "SVA_batchCorrected_sum")
dat_test <- AddMetaData(dat_test, metadata = Matrix::colSums(geomx_counts), col.name = "GeoMx_sum")

# Compare the summed counts for each cell
plot1 <- FeatureScatter(dat_test, feature1 = "SVA_batchCorrected_sum", feature2 = "GeoMx_sum")
plot1
```

==============================================================================
STEP 2: Run Integration on top variance genes
==============================================================================

```{r}
# Set default assay
DefaultAssay(dat) <- "SVA_batchCorrected"

# Split the Seurat object by the block variable
dat.list <- SplitObject(dat, split.by = "Block")

# Normalize data and find variable features for each block separately
dat.list <- lapply(X = dat.list, FUN = function(x) {
  x <- NormalizeData(x)
  x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
  return(x)
})

# Find integration anchors (using top 15 dimensions)
dat.anchors <- FindIntegrationAnchors(object.list = dat.list, dims = 1:15)

# Perform data integration
dat.integrated <- IntegrateData(anchorset = dat.anchors, dims = 1:30, k.weight = 50)

# Set the integrated assay as the default assay
DefaultAssay(dat.integrated) <- "integrated"

# Save integrated object
saveRDS(dat.integrated, file = output_seurat_cca)

# Extract genes from the original and integrated objects
original_genes <- rownames(GetAssayData(dat, layer = "counts"))
integrated_genes <- rownames(GetAssayData(dat.integrated, layer = "data"))

# Find omitted genes
omitted_genes <- setdiff(original_genes, integrated_genes)

# Create a data frame of omitted genes
omitted_genes_df <- data.frame(Gene = omitted_genes)

# Save the omitted genes as a CSV file
write.csv(omitted_genes_df, 
          file = file.path(output_dir_data, "omitted_genes.csv"), 
          row.names = FALSE)

# View the omitted genes data frame
head(omitted_genes_df)
```

==============================================================================
STEP 3: RUN PCA AND UMAP ON CCA-CORRECTED DATA
==============================================================================

```{r}
set.seed(428)

# Scale data and perform PCA for dimensionality reduction
dat.integrated <- ScaleData(dat.integrated)
dat.integrated <- RunPCA(dat.integrated, npcs = 30) 

# Visualize data using UMAP
dat.integrated <- RunUMAP(dat.integrated, dims = 1:8) 

# Find neighbors and clusters for the integrated dataset
dat.integrated <- FindNeighbors(dat.integrated, reduction = "pca", dims = 1:8)
dat.integrated <- FindClusters(dat.integrated, resolution = 1.3)

# Visualize clusters and batch (block variable) on UMAP
p1 <- DimPlot(dat.integrated, reduction = "umap", group.by = "seurat_clusters", label = TRUE)
p2 <- DimPlot(dat.integrated, reduction = "umap", group.by = "Block")
p3 <- DimPlot(dat.integrated, reduction = "umap", group.by = "Group")
p4 <- DimPlot(dat.integrated, reduction = "umap", group.by = "Type")
p5 <- DimPlot(dat.integrated, reduction = "umap", group.by = "Liver_zone")

p1 + p2 + p3 + p4 + p5

```

==============================================================================
STEP 4: UNSUPERVISED CLUSTERING ANALYSIS
==============================================================================


Define Color Palettes
```{r}
# Nature palette for clusters
nature_palette <- c("#4477AA", "#CCBB44", "#228833", "#9970AB", "#BBBBBB")

# Custom color scheme for different metadata variables
custom_colors <- list(
  Treatment = c("infected" = "orange", "uninfected" = "grey23"),
  Group = c("F" = "#941751", "M" = "#005493", "ORX" = "#52B2BF"),
  Liver_zone = c("PP" = "#006400", "PC" = "#91D1C2FF", "IZ" = "#00A087FF"),
  Block = c("1" = "grey10", "2" = "grey40", "3" = "grey80"),
  Type = c("schizont" = "#762A83", "bystander" = "#FFC20A", "mock" = "#BBBBBB"),
  ROI_size = c("200um" = "#290916", "100um" = "#67032f", "75um" = "#a1045a", "50um" = "#BE93D4"),
  
  # Combined phenotype colors
  phenotype_treatmentgroup = c(
    "Py spz & F"   = alpha("#941751", 1),
    "Py spz & M"   = alpha("#005493", 1),
    "Py spz & ORX" = alpha("#52B2BF", 1),
    "Mock & F"     = alpha("#941751", 0.3),
    "Mock & M"     = alpha("#005493", 0.3),
    "Mock & ORX"   = alpha("#52B2BF", 0.3)
  )
)
```

Create Cluster UMAP
```{r}
dat.integrated.clusters <- FindClusters(dat.integrated, resolution = 1.3)


# Relabel clusters: 0->0, 1->2, 2->1, 3->4, 4->3
dat.integrated.clusters@meta.data$seurat_clusters_relabeled <- factor(
  case_when(
    dat.integrated.clusters@meta.data$seurat_clusters == "0" ~ "0",
    dat.integrated.clusters@meta.data$seurat_clusters == "1" ~ "1",
    dat.integrated.clusters@meta.data$seurat_clusters == "2" ~ "2",
    dat.integrated.clusters@meta.data$seurat_clusters == "3" ~ "4",
    dat.integrated.clusters@meta.data$seurat_clusters == "4" ~ "3",
    TRUE ~ as.character(dat.integrated.clusters@meta.data$seurat_clusters)
  ),
  levels = sort(unique(as.numeric(as.character(dat.integrated.clusters@meta.data$seurat_clusters))))
)

# Set the new labels as the active identity
Idents(dat.integrated.clusters) <- dat.integrated.clusters@meta.data$seurat_clusters_relabeled
# Create UMAP with clusters
umap_clusters <- DimPlot(dat.integrated.clusters, 
                         reduction = "umap", 
                         group.by = "seurat_clusters_relabeled", 
                         label = FALSE) +
  scale_color_manual(values = nature_palette) +
  labs(x = "UMAP 1", y = "UMAP 2", title = NULL) +
  coord_fixed(ratio = 1) +
  theme(
    legend.position = "right",
    legend.key.size = unit(0.5, "cm"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    aspect.ratio = 1
  )

ggsave(
  file.path(output_dir_figs, "UMAP_with_clusters_1.3res.pdf"),
  umap_clusters,
  width = 3,
  height = 2.5
)

```

Create Stacked Barplots
```{r}
# Create combined phenotype variable (if not already done)
dat.integrated.clusters@meta.data$phenotype_treatmentgroup <- 
  paste(dat.integrated.clusters@meta.data$Treatment, 
        dat.integrated.clusters@meta.data$Group, 
        sep = "_")

# Add labels
dat.integrated.clusters@meta.data <- dat.integrated.clusters@meta.data %>%
  mutate(
    phenotype_treatmentgroup = case_when(
      phenotype_treatmentgroup == "infected_F" ~ "Py spz & F",
      phenotype_treatmentgroup == "infected_M" ~ "Py spz & M",
      phenotype_treatmentgroup == "infected_ORX" ~ "Py spz & ORX",
      phenotype_treatmentgroup == "uninfected_F" ~ "Mock & F",
      phenotype_treatmentgroup == "uninfected_M" ~ "Mock & M",
      phenotype_treatmentgroup == "uninfected_ORX" ~ "Mock & ORX",
      TRUE ~ phenotype_treatmentgroup
    )
  )

# Define groups of interest
groups_of_interest <- c("phenotype_treatmentgroup", "Liver_zone")

# Initialize list to hold plots
plots <- list()

# Loop over each group of interest
for (group in groups_of_interest) {
  
  if (group %in% colnames(dat.integrated.clusters@meta.data)) {
    
    # Update cluster_data with the current group - USE RELABELED CLUSTERS
    cluster_data <- dat.integrated.clusters@meta.data[, c("seurat_clusters_relabeled", group)]
    colnames(cluster_data) <- c("seurat_clusters", "group_value")
    
    # Calculate counts and proportions by cluster and group
    cluster_data <- cluster_data %>%
      group_by(seurat_clusters, group_value) %>%
      summarise(count = n(), .groups = 'drop') %>%
      group_by(seurat_clusters) %>%
      mutate(proportion = count / sum(count))
    
    # Select the appropriate custom color palette
    color_palette <- custom_colors[[group]]
    
    # Create stacked bar plot
    plot <- ggplot(cluster_data, aes(x = as.factor(seurat_clusters), y = proportion, fill = group_value)) +
      geom_bar(position = "fill", stat = "identity") +
      labs(x = "Clusters", y = "Proportion", fill = group) +
      theme_classic() +
      theme(legend.title = element_blank()) +
      scale_fill_manual(values = color_palette) +
      scale_y_continuous(expand = expansion(mult = c(0, 0)))
    
    plots[[group]] <- plot
  }
}

# Combine plots side by side
combined_plot <- plots[[1]] + plots[[2]] + plot_layout(ncol = 2)

# Save the combined plot
ggsave(
  file.path(output_dir_figs, "Stack_barplot_row_relabeled.pdf"),
  combined_plot,
  width = 6.5,
  height = 3
)
```


Create Heatmap of Top Genes
```{r}
# Define color palette
col_fun <- colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(100)

# Find all markers
dat.markers <- FindAllMarkers(dat.integrated.clusters, only.pos = TRUE)

# Get top 10 genes per cluster
dat.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  slice_head(n = 10) %>%
  ungroup() -> top10

# Create heatmap
heatmap_topgenes <- DoHeatmap(
  dat.integrated.clusters,
  features = top10$gene,
  group.colors = nature_palette
) +
  scale_fill_gradientn(colors = col_fun)

# Option 1: Use cairo_pdf device directly
cairo_pdf(
  file.path(output_dir_figs, "Top_features_by_cluster_1.3res.pdf"),
  width = 5,
  height = 5
)
print(heatmap_topgenes)
dev.off()
```

# ==============================================================================
# STEP 5: PANELS OF UMAPS BY METADATA GROUPS
# ==============================================================================

```{r create-metadata-umaps}
# Ensure we're using the integrated assay and UMAP has been computed
DefaultAssay(dat.integrated) <- "integrated"

# Define custom colors for metadata variables
custom_colors <- list(
  Treatment = c("infected" = "#FA8072", "uninfected" = "#BBBBBB"),
  Block = c("1" = "grey10", "2" = "grey40", "3" = "grey80"),
  Type = c("schizont" = "#762A83", "bystander" = "#FFC20A", "mock" = "#BBBBBB"),
  Group = c("F" = "#941751", "M" = "#005493", "ORX" = "#52B2BF"),
  ROI_size = c("200um" = "#290916", "100um" = "#67032f", "75um" = "#a1045a", "50um" = "#BE93D4"),
  Liver_zone = c("PP" = "#006400", "PC" = "#91D1C2FF", "IZ" = "#00A087FF")
)

# Function to create UMAP plots by metadata
create_umap_plot <- function(data, group_by, title) {
  DimPlot(data, reduction = "umap", group.by = group_by, label = FALSE) +
    scale_color_manual(values = custom_colors[[group_by]]) +
    coord_fixed(ratio = 1) +
    ggtitle(title) +
    labs(x = "UMAP 1", y = "UMAP 2") +
    theme(legend.position = "right", 
          legend.key.size = unit(0.5, "cm"),
      aspect.ratio = 1)  # Additional enforcement of square aspect)
}

# Create UMAP plots for each metadata variable
plots <- list(
  Treatment = create_umap_plot(dat.integrated, "Treatment", "Treatment"),
  Block = create_umap_plot(dat.integrated, "Block", "Block"),
  Type = create_umap_plot(dat.integrated, "Type", "Infection Status"),
  Group = create_umap_plot(dat.integrated, "Group", "Group"),
  ROI_size = create_umap_plot(dat.integrated, "ROI_size", "ROI size"),
  Liver_zone = create_umap_plot(dat.integrated, "Liver_zone", "Liver Zone")
)

# Combine the plots into a grid
combined_plot <- plot_grid(
  plots$Treatment, plots$Block, plots$Type, 
  plots$Group, plots$ROI_size, plots$Liver_zone,
  ncol = 3,
  align = "hv",
  rel_heights = c(1, 1, 1, 1),
  rel_widths = c(1, 1)
)

# Save individual plots as PDFs
for (feature_name in names(plots)) {
  feature_plot <- plots[[feature_name]]
  output_path <- file.path(output_dir_figs, paste0(feature_name, "_umap.pdf"))
  ggsave(
    filename = output_path,
    plot = feature_plot,
    width = 3.7,
    height = 3.7,
    device = "pdf",
    dpi = 300
  )
}

# Save the combined plot as a PDF
combined_output_path <- file.path(output_dir_figs, "All_umaps_colors_by_features.pdf")
ggsave(
  filename = combined_output_path,
  plot = combined_plot,
  width = 12,
  height = 7,
  device = "pdf",
  dpi = 300
)

```

==============================================================================
STEP 6: GENE EXPRESSION UMAPS FOR MANUSCRIPT
==============================================================================

Setup Gene Expression Function
```{r}
# Function to create UMAP plots for genes of interest
create_umap_plot_genes <- function(data, gene_of_interest, title) {
  Seurat::FeaturePlot(data, features = gene_of_interest, pt.size = 1) +
    scale_color_gradientn(
      colors = c("#FFFFBF", "#E08214", "#762A83", "#762A83"),
      values = scales::rescale(c(0, 0.4, 0.8, 1))
    ) +
    coord_fixed(ratio = 1) +
    ggtitle(title) +
    labs(x = "UMAP 1", y = "UMAP 2") +
    theme(
      legend.position = "right",
      legend.key.size = unit(0.5, "cm"),
      aspect.ratio = 1  # Additional enforcement of square aspect
    )
}
```

Figure: Sex at Steady State
```{r}
# Get Group UMAP
Group_umap <- plots[["Group"]]

# Create gene expression plots
Cyp7b1 <- create_umap_plot_genes(dat.integrated, "Cyp7b1", "Cyp7b1")   # Up in M
Cyp2b13 <- create_umap_plot_genes(dat.integrated, "Cyp2b13", "Cyp2b13") # Up in F & ORX

# Make a grid for sex-related genes
Sex_plot <- plot_grid(
  Group_umap, Cyp7b1, Cyp2b13,
  ncol = 3,
  align = "hv",
  rel_heights = c(1, 1, 1, 1),
  rel_widths = c(1, 1)
)

# Save
ggsave(
  filename = file.path(output_dir_figs, "Sex_combined_umap_plot.pdf"),
  plot = Sex_plot,
  width = 10,
  height = 3.5
)


```


Figure: Liver Zonation
```{r}
# Get Liver zone UMAP
Liver_zone_umap <- plots[["Liver_zone"]]

# Create gene expression plots for zonation markers
Cyp2e1 <- create_umap_plot_genes(dat.integrated, "Cyp2e1", "Cyp2e1") # Up in PP
Cyp2f2 <- create_umap_plot_genes(dat.integrated, "Cyp2f2", "Cyp2f2") # Up in PC
Glul <- create_umap_plot_genes(dat.integrated, "Glul", "Glul")       # Up in PP

# Make a grid for zonation genes
Zone_plot <- plot_grid(
  Liver_zone_umap, Cyp2e1, Glul, Cyp2f2,
  ncol = 4,
  align = "hv",
  rel_heights = c(1, 1, 1, 1),
  rel_widths = c(1, 1)
)

# Save
ggsave(
  filename = file.path(output_dir_figs, "Zone_combined_umap_plot.pdf"),
  plot = Zone_plot,
  width = 13,
  height = 3.5,
  dpi = 300
)

```


# ==============================================================================
#  ELBOW PLOT
# ==============================================================================

```{r}

# 1. Use Seurat's ElbowPlot and add a red line at PC 8
p_elbow <- ElbowPlot(dat.integrated, ndims = 30) +
  geom_vline(xintercept = 8, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = 8, y = max(dat.integrated@reductions$pca@stdev[1:30]) * 0.9, 
           label = "PC 8", color = "red", hjust = -0.2, size = 4)

# Save the basic elbow plot
ggsave(here(output_dir_figs, "Supplemental_PC_elbow.pdf"), 
       p_elbow, width = 6, height = 4)


```